---
output:
  md_document:
    includes:
      in_header: ../header.txt
---

### Logistic regression

In this walk-through, you'll learn about the logistic regression model.  Learning goals:  
* Fit a linear probability model and recognize its shortcomings.  
* Fit a logistic regression model.  
* Compute standard errors and confidence intervals using both bootstrapping and the central limit theorem.  
* Make predictions from a logistic regression model.  

Data files:  
* [bballbets.csv](bballbets.csv): Data on college basketball games.  Each row is a game.  The variable we'd like to predict is `homewin`, which is an indicator for whether the home team won the game (1) or lost the game (0).  Our predictor is `spread`, which is the Las Vegas [point spread](http://en.wikipedia.org/wiki/Spread_betting) in favor of the home team for that game.  For example, if the point spread is 10, then the betting markets collectively expect that the home team will win by 10 points.

First we'll load the required libraries and data set.
```{r, message=FALSE}
library(mosaic)
bballbets = read.csv("bballbets.csv", header=TRUE)
```


Let's plot the outcome (homewin) versus the point spread.  Note that, because the homewin variable is either 0 or 1, it is hard to distinguish the individual points.
```{r}
plot(homewin ~ spread, data=bballbets)
```

One simple way around this is to add some artificial jitter to the points, strictly for the sake of visualizing them:
```{r}
plot(jitter(homewin, 0.5) ~ spread, data=bballbets,
  ylab='Home team victory?', xlab='Point spread in favor of home team',
	pch=19, col=rgb(0,0,0,0.2))
```

The points have been jittered up or down enough to distinguish them, but not so much that we can no longer tell which are 0 and which are 1.  Clearly the home team wins more often when there is a big point spread in its favor (and vice versa for the visiting team).

### Calculating empirical frequencies

Our goal is to fit a model that describes how likely the home team is to win, given the Las Vegas point spread.  Here's an especially simple way to proceed: divide the `spread` variable into categories (i.e. discretize it) and compute the fraction of home team wins in each category.  We'll use the `cut' function to divide the point spread at intervals of 5 points:
```{r}
cut_points = seq(-30, 40, by=5)
spread_category = cut(bballbets$spread, cut_points, right=FALSE)
summary(spread_category)
```

Notice that the intervals are closed on the left and open on the right (because we specified `right=FALSE` to the cut function).  Now let's use this categorical variable we've created to calculate the empirical frequency of home-team victories within each bucket:
```{r}
homewin_frequency = mean(homewin ~ spread_category, data=bballbets)
homewin_frequency
```

We can plot the fitted values from this group-wise model as follows:
```{r}
lm0 = lm(homewin ~ spread_category, data=bballbets)
plot(jitter(homewin,0.5) ~ spread, data=bballbets,
  ylab='Home team victory?', xlab='Point spread',
  pch=19, col=rgb(0,0,0,0.2), las=1)
points(fitted(lm0) ~ spread, data=bballbets, col='blue', pch=19)
```

### The linear probability model

The strategy of discretizing the point-spread variable certainly conveys some interesting information, but it's unsatisfying in one important way: it doesn't allow the probability of a home-team win to vary continuously with the point spread.  It treats a point spread of +1 exactly the same as a point spread of +9, which we know can't be right.

We could, of course, discretize the point-spread variable more finely, say into 1-point increments.  But then we'd have very few games falling into each bucket, and our estimate of the empirical win frequency within each bucket would be very noisy.  If you're sharp-eyed, you'll notice from the plot above that this is already a problem even with five-point intervals: the interval [20,24) has a lower empirical win frequency than the interval [15,20), which presumably happened only because of the noise induced by small sample sizes within those intervals.

One alternative approach is called the "linear probability model," in which we fit a linear model for the 0/1 outcome using least squares:
```{r}
lm1 = lm(homewin~spread, data=bballbets)
plot(jitter(homewin,0.5) ~ spread, data=bballbets,
  ylab='Home team victory?', xlab='Point spread',
	pch=19, col=rgb(0,0,0,0.2), las=1)
points(fitted(lm0) ~ spread, data=bballbets, col='blue', pch=19)
abline(lm1)
```

We can interpret the line as the estimated conditional probability of a home-team win, given the Las Vegas point spread.

### Binomial logistic regression

One problem with the straight-line model is that, while it fits the empirical-win frequencies reasonably well towards the middle, it fails badly for extreme values of the point spread.  Notice that the empirical frequencies form something of a foreshortened S shape, and never fall outside the allowed interval [0,1].  By definition, this is impossible to model with a straight line.


A simple fix is to fit a binomial logistic regression model (or a logit model, for short) instead:
```{r}
glm1 = glm(homewin~spread, data=bballbets, family=binomial)
plot(jitter(homewin,0.5) ~ spread, data=bballbets,
  ylab='Home team victory?', xlab='Point spread',
  pch=19, col=rgb(0,0,0,0.2), las=1)
points(fitted(lm0) ~ spread, data=bballbets, col='blue', pch=19)
abline(lm1, col='red')
# Now add the fitted vaues from the logit model
points(fitted(glm1)~spread,data=bballbets, col='green', pch=19)
```

The fitted values from the logit model respect the restriction that probabilities must be between 0 and 1.  They also fit the empirical win frequencies (in blue) much better.


### Quantifying uncertainty in logit models

The coefficients of the logit model can be accessed in the same way as for a linear model:
```{r}
coef(glm1)
```

If we want to compute confidence intervals or standard errors, we know of at least two options.  First, we can bootstrap:
```{r}
boot1 = do(1000)*{
  glm1 = glm(homewin~spread, data=resample(bballbets), family=binomial)
  coef(glm1)
}
hist(boot1$spread)
confint(boot1)
```

We can also appeal to R's summary function, which computes confidence intervals under a normal approximation to the coefficients arising from the central limit theorem:
```{r}
confint(glm1)
```

In this case, the confidence intervals are similar for the two techniques.

### Making predictions from a logit model

Suppose we want to predict the probability of a home-team victory when the point-spread is 5 points in the visiting team's favor (that is, `spread = -5`).  One way to do this is "by hand," using the equation of a logistic regression model:

```{r}
mybeta = coef(glm1)
new_x = -5
psi = mybeta[1] + mybeta[2]*new_x  # linear predictor
predicted_prob = exp(psi)/{1+exp(psi)}
predicted_prob
```

Thus the predicted probability of a home-team victory, given a point spread of 5 points in favor of the visiting team, is about 34%.

We can also use the predict function to streamline this process, which is especially useful if we want to generate predictions at multiple values of x:
```{r}
new_x = data.frame(spread=c(-5, -2, 7, 14))
predict(glm1, newdata=new_x, type='response')
```

We can also use the `predict` function to get the values of the linear predictor (i.e. before it is run through the link function):
```{r}
predict(glm1, newdata=new_x, type='link')
```
