---
output:
  md_document:
    includes:
      in_header: ../header.txt
---

### Gas consumption versus temperature  

In this walk-through, you'll learn how to fit nonlinear models by least squares using a simple trick: adding powers (squared, cubed, etc) of the original predictor variable.  You will also see that new variables can be defined in terms of old ones.

Data files:  
* [utilities.csv](utilities.csv): monthly gas bill versus temperature for a house in Minnesota.  Data source: Daniel T. Kaplan, _Statistical Modeling: A Fresh Approach_, 2009.

First let's load the mosaic library and read in the utilities data set.
```{r, message=FALSE}
library(mosaic)
utilities = read.csv('utilities.csv', header=TRUE)
```

As usual, we'll look at a quick summary of the variables:
```{r}
summary(utilities)
```

### Defining a new variable in terms of existing ones  

Our goal will be to model the monthly gas bill in terms of temperature.  There's a wrinkle here, however: different billing periods have different numbers of billing days:
```{r}
hist(utilities$billingDays, breaks=20)
```

Thus we probably want to be measuring gas usage per day, rather than to the total over each billing period.  Let's define a new variable, called daily.average.gasbill:
```{r}
utilities$daily.average.gasbill = utilities$gasbill/utilities$billingDays
```

Now let's plot the new variable we've created versus temperature, fit a linear model, and add the line to the plot.
```{r}
plot(daily.average.gasbill ~ temp, data=utilities)
lm1=lm(daily.average.gasbill ~ temp, data=utilities)
points(fitted(lm1)~temp, data=utilities, col='red', pch=19)
abline(lm1)
```

This model doesn't do a very good job: we're fitting a linear model to obviously nonlinear data.  You could see this nonlinearity on the original plot, above.  You could also see it on a residual plot, where it's obvious there is still some systematic variation in the residuals as a function of the predictor variable.
```{r}
plot(resid(lm1) ~ temp, data=utilities)
```

### Polynomial regression models  

One approach to address this shortcoming is to fit a parabola: that is, y versus x and x^2.
```{r}
# Fit a model with a quadratic term:
lm2=lm(daily.average.gasbill ~ temp + I(temp^2), data=utilities)
# Replot the data and added the fitted values
plot(daily.average.gasbill ~ temp, data=utilities)
points(fitted(lm2)~temp, data=utilities, col='blue', pch=19)
```

In the model statement, the `I(temp^2)` is the way we tell R to treat temperature-squared as an additional variable in the model.  We could also add higher powers of temperature, although the quadratic fit looks sensible here.

If you want to draw a nice smooth curve, you can plug in the coefficients of the model directly to the `curve` function:
```{r}
plot(daily.average.gasbill ~ temp, data=utilities)
mybeta = coef(lm2)
curve(mybeta[1] + mybeta[2]*x + mybeta[3]*x^2, col='blue', add=TRUE)
```

