---
output:
  md_document
---
  
## Prediction intervals and heteroskedasticity
  
Learning goals:  
- Diagnose heteroskedasticity  
- Fix prediction intervals to account for heteroskedasticity  

Data files:  
* [afc.csv](http://jgscott.github.io/teaching/data/afc.csv): Austin food critics data set.  

```{r, echo=FALSE}
afc = read.csv('afc.csv')
```

After downloading the "afc.csv" file and loading it in, you should first load the mosaic library.  
```{r, message=FALSE}
library(mosaic)
```



```{r}
plot(Price ~ FoodScore, data = afc)
lm1 = lm(Price ~ FoodScore, data = afc)
abline(lm1)
```

```{r}
postdiction = predict(lm1, newdata=afc, interval = 'prediction', level = 0.8)
```

Now let's add the prediction intervals to the plot:
```{r}
plot(Price ~ FoodScore, data = afc)
points(postdiction[,2] ~ FoodScore, data=afc, col='red', pch=19) # lower bound
points(postdiction[,3] ~ FoodScore, data=afc, col='red', pch=19) # upper bound
```

These look really ugly!  They're too wide for low food scores, and too narrow for high food scores.

### Prediction intervals using quantile regression

```{r}
library(quantreg)  # must install first
q10 = rq(Price ~ FoodScore, tau = 0.1, data= afc)
q90 = rq(Price ~ FoodScore, tau = 0.9, data= afc)
```

```{r}
plot(Price ~ FoodScore, data = afc)
points(fitted(q10) ~ FoodScore, data=afc, col='red', pch=19) # lower bound
points(fitted(q90) ~ FoodScore, data=afc, col='red', pch=19) # upper bound
```