---
output:
  md_document:
    includes:
      in_header: ../header.txt
---

### Normal linear regression model

In this walk-through, you'll learn about the normal linear regression model.  Key ideas:  
* Simulating from the normal linear regression model  
* Using R's summary function to extract normal-theory standard errors  
* Checking the validity of the normal-theory assumptions  
* Comparing normal-theory standard errors with bootstrapped standard errors  

Data files:  
* [newspapers.csv](newspapers.csv): Sunday and weekly circulation numbers for major metropolitan newspapers.    


### Simulating from the model

The normal linear regression model assumes that the residuals in a regression follow a normal distribution.  To get a sense of what this means, it helps to simulate some data for which this assumption is true by construction.

```{r, message=FALSE}
# Load the mosaic library
library(mosaic)
```

```{r}
# Choose some true parameters
beta0 = -2
beta1 = 1.25
sigma = 3
# Simulate some predictors between 0 and 10
n = 50
x = runif(n, 0, 10)
# Simulate the residuals and form the responses
epsilon = rnorm(n, 0, sigma) # mean of zero, std. dev. of sigma
y = beta0 + beta1*x + epsilon
plot(x,y, pch=19)
# Plot the true line
abline(beta0,beta1, col='red')
```

Try executing the block of code several times to get a sense of how the data change from run to run.  In each case, the simulated data look a lot like what we might expect to see in a "real" data set.

A natural next step to build our intuition is to run a Monte Carlo simulation.  Each simulated data set will have the same underlying parameters, just a different set of residuals simulated from a normal distribution.

```{r}
mc1 = do(100)*{
  # Simulate new residuals each time
  epsilon = rnorm(n, 0, sigma)
  y = beta0 + beta1*x + epsilon
  lm_samp = lm(y~x)
  coef(lm_samp)
}
# Inspect and summarize the sampling distributions
hist(mc1$x)
hist(mc1$Intercept)
sd(mc1$x)
sd(mc1$Intercept)
```

### Using normality assumptions to quantify uncertainty

For real data, the assumption that residuals are normally distributed can be either a good approximation or a bad approximation to reality.  When it's a good approximation, the nice result is that we have simple mathematical formulas for the standard error of the sampling distribution of the least-squares estimator.  We need not appeal to these formulas explicitly, because they are embedded into all common statistical software, including R.  They can be applied to the data set as long as we know the right commands.  Let's see how this works.

First we'll try this on simulated data, where we can compare the normal-theory answers to Monte Carlo simulations.  Let's start by simulating a single data set from the normal linear regression model:
```{r}
epsilon = rnorm(n, 0, sigma)
y = beta0 + beta1*x + epsilon
lm_samp = lm(y~x)
plot(x,y, pch=19)
abline(lm_samp, col='red')
summary(lm_samp)
```

Notice that R's summary function has a column called "standard error."  In fact, these are precisely the standard errors that are calculated using the formulas for the normal linear regression model.  (These formulas are given in any standard textbook on regression analysis, including in our own course packet.)

To provide a sanity check on these numbers, let's estimate the standard error by Monte Carlo simulation under the normal model.  We will re-run the simulation above with many more than 100 Monte Carlo samples:
```{r}
mc1 = do(5000)*{
  # Simulate new residuals each time
  epsilon = rnorm(n, 0, sigma)
  y = beta0 + beta1*x + epsilon
  lm_samp = lm(y~x)
  coef(lm_samp)
}
sd(mc1$x)
sd(mc1$Intercept)
```
These numbers should be pretty close to those given by the summary function.  It looks like the normal-theory standard errors are in the ballpark.

Now let's take a look at some real data and apply the same logic.  First load the newspapers data set:
```{r}
newspapers = read.csv('newspapers.csv')
```

The two variables and the weekday and Sunday circulation numbers (in thousands of papers) for 34 major metropolitan dailies.  Let's fit a model that uses daily circulation to predict Sunday circulation:
```{r}
plot(Sunday ~ Daily, data=newspapers)
lm1 = lm(Sunday ~ Daily, data=newspapers)
abline(lm1)
```

If we ask for a summary of the model, we will get the normal-theory standard errors:
```{r}
summary(lm1)
```

We can also get confidence intervals under the normal-theory assumption:
```{r}
confint(lm1, level = 0.95)
```

Finally, we can also use the model to get prediction intervals under the normal-theory assumptions.
```{r}
new_data = data.frame(Daily = c(250, 500, 750))
predict(lm1, new_data, interval='prediction', level = 0.95)
```
Importantly, this prediction interval "bakes in" uncertainty about the parameters of true regression line.  (The formulas for doing so are all taken care of "under the hood.")  It is therefore slightly wider than a naive prediction interval, in which parameter uncertainty is ignored.



### Residual diagnostics

Of course, if you're going to apply the normal-theory formulas, it's always a good idea to check whether the assumptions look reasonable.  These assumptions are:  
1) The residuals are normally distributed.  
2) The residuals are independent of each other.  
3) The variance of the residuals is constant across the whole rage of the predictor (x) variable.   

We can check these assumptions visually.
```{r}
hist(resid(lm1), 20)
```

There are only 34 residuals, and so they don't form an especially pretty bell curve.  But they at least don't look wildly divergent from a normal distribution.  To check whether the residuals look correlated with one another, or whether their variance changes noticeably as a function of the predictor variable, we can plot the residuals versus Daily circulation.
```{r}
plot(resid(lm1) ~ Daily, data = newspapers)
```

There are no obvious red flags here.  If you look for awhile at the figure, you might argue that the residuals are bit more bunched up for small values of daily circulation (around 200), and a bit more spread out for large values.  But if the effect is there at all, it certainly isn't dramatic.  These plots lend further credence to the standard errors calculated from the normality assumption.

### Comparison with bootstrapped standard errors  

As a further sanity check, let's compare these normal-theory standard errors to those we get from bootstrapping.
```{r}
my_boot = do(1000)*{
  lm_boot = lm(Sunday ~ Daily, data=resample(newspapers))
  coef(lm_boot)
}
sd(my_boot$Intercept)
sd(my_boot$Daily)
```
These should be reasonably close to the normal-theory standard errors.
 
Note: if you want to extract the standard errors alone, rather than view them in a table, you can do so like this:
```{r}
summary(lm1)$coefficients[,2]
```
