---
output:
  md_document:
    includes:
      in_header: ../header.txt
---

## Manufacturing flaws in circuit boards

In this case study, you'll look at data from a quality-control experiment from AT&T's process for manufacturing printed circuit boards.  The data are in [solder.csv](solder.csv), and come to us courtesy of the R package `faraway.`

```{r, echo=FALSE}
solder = read.csv('solder.csv')
```

Load the data set using the Import Dataset button.  Then load the mosaic library and begin by summarizing the data set.
```{r, message=FALSE}
library(mosaic)
summary(solder)
```

The response variable, skips, is the number of solder skips on the circuit board (i.e. manufacturing flaws) that are apparent from a visual inspection.  The remaining variables refect different choices in the manufacturing process.  The goal is to understand which combination of choices leads to the lowest number of skips, and therefore the most reliable manufacturing process.  We'll focus on three such variables:  
* Opening: the size of the opening on the solder gun (small, medium, or large)  
* Solder: the thickness of the alloy used for soldering (thick or thin)  
* Mask: according to Wikipedia, a solder mask is "a thin lacquer-like layer of polymer that is usually applied to the copper traces of a printed circuit board (PCB) for protection against oxidation and to prevent solder bridges from forming between closely spaced solder pads."  This experiment involved five different masks, labeled as A1.5, A3, A6, B3, and B6.


Let's start by looking at how the number of skips depends on each variable individually.
```{r}
bwplot(skips ~ Solder, data=solder)
bwplot(skips ~ Opening, data=solder)
bwplot(skips ~ Mask, data=solder)
```

Each seems to predict some of the variation in solder skips.

Build a model to predict solder skips using these three predictor variables.  Include whatever combination of main effects and interaction terms you deem appropriate.  Justify your final choice of model and explain what conclusions you can draw from your analysis of this data set.

Citation for original data source: Comizzoli, R. B., J. M. Landwehr, and J. D. Sinclair (1990). Robust materials and processes: Key to reliability. AT&T Technical Journal 69(6), 113-128.

