---
output:
  md_document:
    includes:
      in_header: ../header.txt
---

## Test scores and GPA for UT graduates

In this walk-through, you'll learn how to summarize and visualize the following kinds of relationships:   
- between a numerical variable and a categorical variable, via group-wise means and boxplots.  
- between two numerical variables, via scatter plots and correlation coefficients.   
- three variables, using lattice plots.  

You will also learn how to change more of the default plot settings in R plots.

You'll need this data file:  
* [ut2000.csv](ut2000.csv): data on SAT scores and graduating GPA for every student who entered the University of Texas at Austin in the fall of 2000 and went on to graduate within 6 years.

### Preliminaries

First, in order to complete this analysis we need to load in the `mosaic` library from within RStudio:  
```{r, message=FALSE}
library(mosaic)
```

Next, download the ut2000.csv file and read it in.
```{r}
ut2000 = read.csv('ut2000.csv')
summary(ut2000)
```

### Between-group and within-group variation

School (which of the 10 different undergraduate schools at UT the student graduated from) is a natural grouping variable here.  To see both the between-group and within-group variation, let's examine a boxplot of SAT math scores (SAT.Q) versus school.
```{r, fig.width=9}
bwplot(SAT.Q ~ School, data=ut2000, main="SAT Math Scores by College")
```

The `bwplot` command is available because we loaded the `mosaic` package (if you didn't do this, the bwplot command will produce only an error). The aesthetics of `bwplot` are a little nicer than the basic R command `boxplot`, which would also work just fine.  You may also notice the names of the colleges along the x axis running together, which you can fix by clicking "Zoom" in the Plots tab and manually resizing the window.

The boxplot allows you to get a sense of whether the between-group or within-group variation of SAT math scores is larger.  
- Between-group variation: how much do the central dots for each group differ from one another?  
- Within-group variation: how spread out are the cases within each group?  

Let's first focus the between-group variation, by examining how the group means differ from one another.
```{r}
mean(SAT.Q ~ School, data=ut2000)
```
Informally, it looks like the typical difference between group means is about 40 points.

We can also examine the within-group variation, by looking at the standard deviation of SAT math scores within each college:
```{r}
sd(SAT.Q ~ School, data=ut2000)
```
It looks as if SAT math scores vary most within Communications, and least within Social Work.  

The `favstats` command will compute both the mean and standard deviation, along with several of your favorite statistics (like the minimum/maximum values, the first and third quartiles, and the median), for each college:
```{r}
favstats(SAT.Q ~ School, data=ut2000)
```


### Two numerical variables  

Our basic tool for visualizing the relationship between two numerical variables is the scatter plot.  Let's make one which shows graduating GPA versus SAT math scores for all students.
```{r}
plot(GPA ~ SAT.Q, data=ut2000)
```

If you want to get fancy, you can change the type of mark (pch), size (cex), or color (col) of the plotted points:
```{r}
plot(GPA ~ SAT.Q, data=ut2000, pch=19, cex=0.5, col='grey')
```
Try `?plot`, `?points`, and `?par` to see some more of the plotting options.

To compute a correlation coefficient, use the `cor` function:
```{r}
cor(GPA ~ SAT.Q, data=ut2000)
```

### Scatter plot matrix 

To visualize the relationships among three or more numerical variables, we can create a pairs plot showing each bivariate relationship.  In this case, we'll make a matrix of scatter plots corresponding to the 1st, 2nd, and 5th columns of the `ut2000` data set (for SAT Verbal, SAT Math, and GPA, respectively):
```{r}
pairs(ut2000[,c(1,2,5)])
```
The brackets to the right of our data set, `ut2000[,c(1,2,5)]`, say "give me all rows from the 1st, 2nd, and 5th columns."  If you wanted, say, the first 100 rows of these columns, you'd use `ut2000[1:100,c(1,2,5)]`.  But because we left the row index blank in the command above, R gives us all the rows by default.

You might find this pairs plot redundant---for example, the plot of SAT.V versus GPA (row 1, column 3) contains the same information as the plot of GPA versus SAT.V (row 3, column 1).  We've just flipped which variable appears on the vertical axis. To suppress this redundancy, pass in the following flag:
```{r}
pairs(ut2000[,c(1,2,5)], upper.panel=NULL)
```


### Lattice plots  

Finally, if we want to see this bivariate relationship plotted separately for each of the 10 colleges, we need what's called a lattice plot.  A lattice plot allows us to see whether, and how, the relationship between two variables is modulated by a third variable.  The `xyplot` command produces one:
```{r}
xyplot(GPA ~ SAT.Q | School, data=ut2000)
```
The vertical bar (|) should be read as "conditional upon" or "stratified by."    In this case, the GPA versus SAT.Q relationship looks broadly similar across all colleges.

We can also make a lattice of boxplots.  For example, we might be interested the relationship between graduating GPA and college depends on a student's incoming SAT score.  We can do this by discretizing SAT score into intervals.  (You'll remember we did this using the `cut` function in an [earlier walkthrough](http://jgscott.github.io/teaching/r/titanic/titanic.html).)  We can then use these intervals to define a lattice plot.  

```{r}
ut2000$SATcat = cut(ut2000$SAT.C, breaks=c(0, 1000, 1200, 1400, 1600))
bwplot(GPA ~ School | SATcat, data=ut2000)
```

This is really ugly because all the axis labels are running together.  We can rotate the college labels to 45 degrees using the (admittedly also ugly) command below:
```{r}
bwplot(GPA ~ School | SATcat, data=ut2000, scales=list(x=list(rot=45)))
```
