---
output:
  md_document
---

In this walk-through, you'll learn how to measure and visualize dispersion of a single quantitative variable.  You will also learn how to change some of the default plot settings in R, like changing the axis labels or the number of breaks in a histogram.  

Data files:  
* [citytemps.csv](http://jgscott.github.io/teaching/data/citytemps.csv): daily average temperatures in San Diego, CA and Rapid City, SD.  

### Preliminary steps

First, load the mosaic library, like this.
```{r, message=FALSE}
library(mosaic)
```
There will be a lot of information returned to the console when you load the library, but you only have to worry if you see an actual error, like this:
```
Error in library(mosaic) : there is no package called ‘mosaic’
```
This means you haven't installed the package!  You'll only have to do the installation once, but you will have to load the mosaic library (as above) at the beginning of each new R session if you intend to use it.  If you haven't installed the mosaic library, see the [instructions here.](../basics/installing_library.md)

The mosaic library has two useful functions, `pdata` and `qdata`, for dealing with quantiles of data sets.  If you don't load the library, some of the commands below won't work.

Next, download the citytemps.csv file and read it in. You can use RStudio's Import Dataset button, or use the read.csv command:
```{r}
citytemps = read.csv('citytemps.csv')
```
If you do this second option, make sure you're working from the directory where the data lives, or put the full path to the file between the quotation marks.


### A simple histogram

We'll start off by getting R to summarize the variables:
```{r}
summary(citytemps)
```

You can see we have 17 years of temperature data on San Diego and Rapid City.  Let's focus on San Diego. We'll start with a histogram, one of the most basic plots for a single quantitative variable:

```{r}
hist(citytemps$Temp.SanDiego)
```

If you want the histogram to have more bins, you can specify this via the `breaks` flag:
```{r}
hist(citytemps$Temp.SanDiego, breaks=30)
```
This is called "adding a flag", (in this case, `breaks=30`) to the basic command.  This is a very common paradigm in R: a command will have a default behavior which you can modify by providing a function with optional flags.  In this case, we're asking R to provide a histogram with (approximately) 30 breaks.

R plots can have a lot more bells and whistles than this, some of which we'll encounter below.  To give a simple example, we can add a vertical line at the sample mean.
```{r}
muSanDiego = mean(citytemps$Temp.SanDiego)
hist(citytemps$Temp.SanDiego, breaks=30)
abline(v=muSanDiego, col='red')
```

### Measuring dispersion: standard deviation and quantiles  

You already know one way of summarizing the dispersion of this distribution: the standard deviation.
```{r}
sd(citytemps$Temp.SanDiego)
```

Another measure of dispersion is the coverage interval: that is, an interval covering a specified fraction of the observations.  For example, to get a central 50% coverage interval, we'd need the 25th and 75 percentiles of the distribution.  By definition, 50% of the observations are between these two numbers.  You can get these from the `qdata` function.
```{r}
qdata(citytemps$Temp.SanDiego)
```
This tells you that, for example, the 75th percentile of the data is at 67.3 degrees.

You can also ask for different quantiles by passing in a flag called "p" (for probability) like this:
```{r}
qdata(citytemps$Temp.SanDiego, p=c(0.05, 0.95))
```
In this case, `p=c(0.05, 0.95)` is flag that modifies the default behavior of the quantile function to yield the requested quantiles, in this case the 5th and 95th percentiles.  We could use these numbers to construct a 90% coverage interval.

We can also do the inverse calculation using the `pdata` function, asking which quantile a specific value corresponds to:
```{r}
pdata(citytemps$Temp.SanDiego, q=c(60, 70))
```
This tells us that 60 degrees is the 33rd percentile of the data set and that 70 degrees is the 88th percentile.

To summarize:  
- You give `qdata` a data set and a requested quantile, and it returns the corresponding value at that quantile.  
- You give `pdata` a data set and a value, and it returns the corresponding quantile for that value.  


### Standardizing using z-scores

Which temperature is more extreme: 50 degrees in San Diego, or 10 degrees in Rapid City?  In an absolute sense, of course 10 degrees is a more extreme temperature.  But what about in a relative sense?  In other words, is a 10-degree day more extreme _for Rapid City_ than a 50-degree day is for San Diego?

This question could certainly be answered using quantiles, which you've already learned how to handle.  But let's discuss a second way: by calculating a z-score for each temperature.  

A z-score is the number of standard deviations by which some observation is above the mean.  (So if a z-score is negative, then the corresponding observation is _below_ the mean.)  To calculate a z-score, we subtract the mean and divide by the standard deviation.  For a 50-degree day in San Diego, this is:
```{r}
(50 - mean(citytemps$Temp.SanDiego)) / sd(citytemps$Temp.SanDiego)
```
Or about 2.3 standard deviations _below_ the mean.

On the other hand, for a 10-degree day in Rapid City, this is
```{r}
(10 - mean(citytemps$Temp.RapidCity)) / sd(citytemps$Temp.RapidCity)
```
Or about 1.9 standard deviations below the mean.  Thus a 50-degree day in San Diego is actually more extreme than a 10-degree day in Rapid City!

As this example suggests, z-scores are useful for comparing numbers that come from different distributions, with different statistical properties. It tells you how extreme a number is, relative to other numbers from that some distribution.

### Fancier histograms

To close, we'll make a histogram for Rapid City, except we'll change the default title and x-axis label to something a bit more fit for human consumption.
```{r}
hist(citytemps$Temp.RapidCity, main="Average Daily Temperatures in Rapid City, 1995-2011", xlab='Temperature')
```

Notice again that we've used optional flags (main and xlab) to change the main title and x-axis label, respectively.

Next, let's stack the two histograms on top of each other, to make a multi-frame plot.  We do this with the following code block.  The first line says to make a multi-frame plot (filled in along the rows) with 2 rows and 1 column.  The next two plotting commands then fill in the two frames.

```{r, fig.height=8}
par(mfrow=c(2,1))
hist(citytemps$Temp.SanDiego)
hist(citytemps$Temp.RapidCity)
```

This won't do: notice that the axes and bin sizes differ between the two plots.  This makes it hard to compare the two distributions at a glance.  We need to align these two plots to have the same axes and bins.  Just as we did above, we'll do this by passing additional flags to the `hist` function.

First, we must define a set of breakpoints for the histogram grams.  We'll do this with the `seq` (which stands for sequence) command:
```{r}
mybreaks = seq(-20, 92, by=2)
```
This says to make a sequence running from -20 to 92 degrees in increments of 2.  You'll see the whole sequence if you type `mybreaks` directly into the console"
```{r}
mybreaks
```

Now we can make the histograms using these custom bins.  We'll also change the x and y axes using the xlim and ylim arguments:
```{r, fig.height=8}
par(mfrow=c(2,1))
hist(citytemps$Temp.SanDiego, breaks=mybreaks, xlim=c(-20,100), ylim=c(0, 760))
hist(citytemps$Temp.RapidCity, breaks=mybreaks, xlim=c(-20,100), ylim=c(0, 760))
```
Now the bins and axes are comparable, making the distributions themselves much easier to compare.

In the expression above, you'll notice that we used the same `breaks` flag as before, except this time we passed in a full set of breaks rather than an integer.  R lets us do either.


### Optional 

Below, I've included a set of commands that produce a much prettier plot.  Feel free to go line by line and modify these commands to see what behavior they produce.  You can learn a lot about R this way.

```{r}
mybreaks = seq(-20, 92, by=2)
par(mfrow=c(1,1), mar=c(3,0,1,3), mgp=c(2,1,0))
hist(citytemps$Temp.SanDiego, breaks=mybreaks, xlab="Average Daily Temperature (F)", main="", border="darkgrey", col="grey", axes=FALSE, ylim=c(0, 760))
hist(citytemps$Temp.RapidCity,breaks=mybreaks,add=TRUE, border=rgb(0,100,0,100,maxColorValue=255), col= rgb(0,100,0,50,maxColorValue=255))
axis(4,at=seq(0,700,by=100), las=1,tick=FALSE)
axis(1,pos=0)
text(55, 770, "San Diego, CA", pos=4, font=2)
text(30, 260, "Rapid City, SD", pos=4, font=2)
```

Remember you that can always use the question mark, e.g.
```{r}
?text
```
to see the list of available arguments/flags for a given function.  
