---
output:
  md_document
---

## Simulating stocks and bonds under a bivariate normal

In this case study, you'll use historical data on stock and bond returns, in order to simulate the performance of a portfolio under various assumptions about the correlation between these two assets.

### Preliminaries

First load the `mosaic` and `mvtnorm` libraries.
```{r, message=FALSE}
library(mosaic)
library(mvtnorm)
```

Next, read in the function from my website that allows easy simulation under the bivariate normal model:
```{r}
source("http://jgscott.github.io/teaching/r/mvnorm/rbvnorm.R")
```

## Simulating portfolios

There are three key ingredients for simulating a portfolio of stocks and bonds (specifically, [ten-year U.S. treasury bonds](http://www.investopedia.com/terms/t/treasurybond.asp), although the same idea works for any asset class.)  

First, we need some statistical assumptions about the performance of these assets.  Let's start with these:  
- stocks have a long-term average net return of 6.5%, net of inflation, with a standard deviation of 19.5%.  
- 10-year treasury bonds have a long-term average net return of 1.7%, with a standard deviation of 7.5%.  
- the correlation between stock and bond returns in any given year is -0.15.  

Second, we need a probability model based on these statistical assumptions that allows us to simulate from the joint distribution of stocks and bonds.  A natural candidate is the bivariate normal.  Here are the parameters we'll use:  
```{r}
mu_stocks = 0.065
mu_bonds = 0.017
sd_stocks = 0.195
sd_bonds = 0.075
rho = -0.15 
```

As a reminder, we call the `rbvnorm` function with these parameters in order to simulate random draws from this joint distribution.  For example, here's where 50 years of returns simulated from these model looks like: 
```{r}
returns = rbvnorm(50, mu_stocks, mu_bonds, sd_stocks, sd_bonds, rho)
plot(returns)
```

Finally, we need a way of chaining together our investment performance across the years, to simulate a single imaginary future.  We do this using a for loop.  For example, the following loop allowed us to simulate one 40-year investment trajectory of a portfolio with stocks only, under a (one-variable) normal model:  
```{r}
Wealth = 10000
Horizon = 40
for(year in 1:Horizon) {
  return_stocks = rnorm(1, mu_stocks, sd_stocks)
	Wealth = Wealth * (1 + return_stocks)
}
Wealth
```

We can easily extend this idea to a bivariate normal model for stocks and bonds, tracking our wealth in each asset at each step.  Here we assume an 60/40 allocation of wealth to stocks and bonds, respectively:   
```{r}
total_wealth = 10000
weights = c(0.6, 0.4) # how much of your wealth in each asset?
wealth_by_asset = total_wealth * weights
Horizon = 40
for(year in 1:Horizon) {
  # Simulate a bivariate normal set of returns
  returns = rbvnorm(1, mu_stocks, mu_bonds, sd_stocks, sd_bonds, rho)
  # Update wealth in each asset
	wealth_by_asset = wealth_by_asset * (1 + returns)
	# rebalance
	total_wealth = sum(wealth_by_asset)
	wealth_by_asset = total_wealth * weights
}
total_wealth
```


### The case study  

The goal of this case study is simple: to understand how the correlation between stocks and bonds affects the long-term performance of your investment portfolio.  After all, this correlation isn't constant.  Rather, [it changes over time in response to larger macro-economic forces.](https://www.rba.gov.au/publications/bulletin/2014/sep/pdf/bu-0914-8.pdf)    

Suppose that you start with $10,000, and that you allocate 60% of your money to stocks (with the remaining 40% to bonds).  Use Monte Carlo simulation to answer the following questions:  
1. What are the mean and standard deviation of simulated wealth after 40 years, assuming that stocks and bonds have a correlation of -0.3?  
2. What if stocks and bonds instead have a positive correlation of 0.3?  

In each case, assume that you rebalance your portfolio every year to retain the target 60/40 allocation.  

The same script that you build for this case study can be used to investigate any set of assumptions about the statistical properties of future stock/bond returns.  


## Optional topic: the statistical properties of stocks and bonds   

### The data  

In this (optional) section, I'll walk through the process of estimating historical returns for stocks and bonds.  We'll use the following dat:  
[stocks_bonds.csv](http://jgscott.github.io/teaching/data/stocks_bonds.csv): historical returns (1928 - 2016) of the S&P 500 stock index and 10-year treasury bonds.  

```{r, echo=FALSE}
stocks_bonds = read.csv('http://jgscott.github.io/teaching/data/stocks_bonds.csv')
```

Download this data and read it in, using the Import Dataset button. 

Let's make histograms of these data here, just to get a sense of what we're working with:  
```{r}
hist(stocks_bonds$SP500)
hist(stocks_bonds$TBOND10)
```

### Estimating the long-term mean and standard deviations    

Let's calculate the historical average returns for stocks and 10-year treasury bonds.  Because we're talking about financial returns here, the most appropriate measure of "average" is the [geometric mean](http://www.investopedia.com/articles/investing/071113/breaking-down-geometric-mean.asp), rather than the more usual arithmetic mean:
```
mu_stocks_gross = exp(mean(log(1+stocks_bonds$SP500))) - 1
mu_bonds_gross = exp(mean(log(1+stocks_bonds$TBOND10))) - 1
```

These are gross returns, without accounting for inflation.  To get a net average return, we'll subtract the long-term average inflation rate of 3.2%:
```
mu_stocks = mu_stocks_gross - 0.032
mu_bonds = mu_bonds_gross - 0.032
```

Calculating the standard deviations is a bit easier:
```{r}
apply(stocks_bonds, 2, sd)
```



