---
output:
  md_document
---

## Optimizing a function in R

You'll often find yourself wanting to optimize some function, to find either a minimum or maximum value.  Let's see how to do this in R.  You'll need the mosaic library, so make sure to load it first: 
```{r, message=FALSE}
library(mosaic)
```

### Plot and point  

As a simple case, we'll consider the function $f(x) = x^{-2.5} (x - 1)$.   Let's first plot this function:
```{r}
curve(x^(-2.5) * (x-1))
```

Why is this so ugly?  It's because we're plotting this function over the default domain from the `curve` function, which is [0,1].  Let's change the domain to be plotted:
```{r}
curve(x^(-2.5) * (x-1), from = 1, to =5)
```

That's better!  We can already see that the maximum is somewhere around 1.7.  We can even zoom in further on this region:
```{r}
curve(x^(-2.5) * (x-1), from = 1.6, to = 1.8)
```

Just plotting and pointing ("hey, here's the optimum!") is enough to tell us that this function obtains a maximum around $x=1.67$.

### Using `makeFun` and `optimize`

As an alternative, you can follow a simple two-step process to get R to optimize a function for you -- that is, without plotting and pointing.

First, you define your own function, using the mosaic library's `makeFun` function.  It works like this:
```{r, message=FALSE}
myfunction = makeFun(x^(-2.5) * (x-1) ~ x)
```
On the left-hand side of the tilde (~) is the function f(x), and on the right-hand side is that function's variable (x).  This step creates a _user-defined function_ for R to optimize.  

The second step is to call the `optimize` function on your own user-defined function.  The following code snippet says to find a maximum of `myfunction` over the interval [0,10].  
```{r}
optimize(myfunction, lower = 0, upper = 10, maximum = TRUE)
```
The output tells you that the value of $x$ at which your function obtains a maximum is 1.6666 (rounded off), and that the value at that point is $f(x) = 0.1859$.

Three notes here: 

1) You can call your user-defined function just like any other function in R.  For example:
```{r}
myfunction(4)
```
This tells you that the value your function takes at x=4 is f(x) = 0.09375.

2) If you wanted to find the minimum, you would chaince "maximum = TRUE" to "maximum = FALSE".  (Of course, this function is unbounded below, so it has no minimum.)  

3) In the above code, you are passing one function (`myfunction`) as an argment to another function (`optimize`).  This might make your head spin a bit at first, but it's a perfectly valid thing to do.  Many functions in R, in fact, _expect_ that their arguments are other functions.  `curve` is an example:
```{r}
curve(myfunction, from=1, to=3)
```

### A second example

Just to put all this together, let's maximize a different function, $f(x) = exp(-0.5*(x-3)^2)$.  

```{r}
myotherfunction = makeFun(exp(-0.5*(x-3)^2) ~ x)
optimize(myotherfunction, lower = -10, upper = 10, maximum=TRUE)
```
This tells us that the maximum occurs at x=3, which we can verify by plotting:
```{r}
curve(myotherfunction, from=0, to=6)
```

