---
output:
  md_document
---

## Optimizing a function in R

You'll often find yourself wanting to optimize some function, to find either a minimum or maximum value.  Let's see how to do this in R.  You'll need the mosaic library, so make sure to load it first: 
```{r, message=FALSE}
library(mosaic)
```

### Plot and point  

As a simple case, we'll consider the function $f(x) = x^{-2.5} (x - 1)$.   Let's first plot this function:
```{r}
curve(x^(-2.5) * (x-1))
```

Why is this so ugly?  It's because we're plotting this function over the default domain from the `curve` function, which is [0,1].  Let's change the domain to be plotted:
```{r}
curve(x^(-2.5) * (x-1), from = 1, to =5)
```

That's better!  We can already see that the maximum is somewhere around 1.7.  We can even zoom in further on this region:
```{r}
curve(x^(-2.5) * (x-1), from = 1.6, to = 1.8)
```

Just plotting and pointing ("hey, here's the optimum!") is enough to tell us that this function obtains a maximum around $x=1.67$.

### Using `makeFun` and `optimize`

As an alternative, you can follow a simple two-step process to get R to optimize a function for you -- that is, without plotting and pointing.

First, you define your own function, using the mosaic library's `makeFun` function.  It works like this:
```{r, message=FALSE}
myfunction = makeFun(x^(-2.5) * (x-1) ~ x)
```
On the left-hand side of the tilde (~) is the function f(x), and on the right-hand side is that function's variable (x).  This step creates a _user-defined function_ for R to optimize.  

The second step is to call the `optimize` function on your own user-defined function.  The following code snippet says to find a maximum of `myfunction` over the interval [0,10].  
```{r}
optimize(myfunction, lower = 0, upper = 10, maximum = TRUE)
```
The output tells you that the value of $x$ at which your function obtains a maximum is 1.6666 (rounded off), and that the value at that point is $f(x) = 0.1859$.

Three notes here: 

First, you can call your user-defined function just like any other function in R.  For example:
```{r}
myfunction(4)
```
This tells you that the value your function takes at x=4 is f(x) = 0.09375.

Second, if you wanted to find the minimum, you would change "maximum = TRUE" to "maximum = FALSE" in the above call to optimize.  (Of course, the function we chose is unbounded below, so it has no minimum.)  

Third, in the above code, you are passing one function (`myfunction`) as an argment to another function (`optimize`).  This might make your head spin a bit at first, but it's a perfectly valid thing to do.  Many functions in R, in fact, _expect_ that their arguments are other functions.  `curve` is an example:
```{r}
curve(myfunction, from=1, to=3)
```

### A second example

Just to put all this together, let's maximize a different function, $f(x) = exp(-0.5*(x-3)^2)$.  

```{r}
myotherfunction = makeFun(exp(-0.5*(x-3)^2) ~ x)
optimize(myotherfunction, lower = -10, upper = 10, maximum=TRUE)
```
This tells us that the maximum occurs at x=3 with value f(x) = 1, which we can verify by plotting:
```{r}
curve(myotherfunction, from=0, to=6)
```

### Without the mosaic package

The `makeFun` command from the mosaic package is pretty intuitive for defining simple functions.  But you can do the same thing in "base" R.  Here's how to define your own function using `function` command:   

```{r, message=FALSE}
myfunction = function(x) {
  y = x^(-2.5) * (x-1)
  return(y)
}
```

This is a little bit more verbose way of defining your own function, but it's also more flexible.  The basic structure is:   
1. Put the function's variables inside the parentheses of the `function` definition, e.g.  `myfunction = function(x)` here.  These are called the arguments of the function.  Here is there is only one argument, called x.  
2. Put executable commands between the braces {}.  E.g. here we define a new variable y, calculated from the argument (input) x that is passed to the function whenever it gets called.  Here the commands to be executed are really short, but these can be as long or complicated as you want.  
3. The last line of the function should return whatever you want the function to return---in this case, the variable y, representing the value of the function at the input x.  

Once we've done this, we have a callable function, e.g.
```{r}
myfunction(2)
myfunction(3.15)
```

And `optimize` works exactly the same as it did before.  
```{r}
optimize(myfunction, lower = 0, upper = 10, maximum = TRUE)
```
