---
output:
  md_document:
    includes:
      in_header: ../header.txt
---

### Tests of association in 2x2 tables

In this walk-through, you will revisit the Titanic data to learn about permutation tests.  Key concepts and skills:  
* test statistic  
* relative risk  
* odds ratio  
* permutation test  

Data files:  
* [TitanicSurvival.csv](TitanicSurvival.csv)  

First download the TitanicSurvival.csv file and read it in. You can use RStudio's Import Dataset button, or the read.csv command:
```{r, message=FALSE}
library(mosaic)
TitanicSurvival = read.csv('TitanicSurvival.csv')
```

Recall that this data set shows the name of each passenger and whether they survived, along with their age, sex, and cabin class.
```{r}
head(TitanicSurvival)
```

### Relative risk in 2x2 tables

One of the very first contingency tables we made looked at survival status stratified by sex:
```{r}
t1 = xtabs(~sex + survived, data=TitanicSurvival)
prop.table(t1, margin=1)
```

This seems to suggest a strong association between survival status and sex. A natural _test statistic_ to quantify this association between the rows and columns of this table is the [relative risk](http://en.wikipedia.org/wiki/Relative_risk) of dying: that is, the probability of dying for men, divided by the probability of dying for women.  We can compute this as follows:
```{r}
# Save the table output in an object called t1
t1 = xtabs(~sex + survived, data=TitanicSurvival)
# Convert to proportions normalized by row
p1 = prop.table(t1, margin=1)
# Calculate the relative risk of dying for both men and women
# in terms of the individual cells of the table
risk_female = p1[1,1]
risk_male = p1[2,1]
relative_risk = risk_male/risk_female
relative_risk
```

The relative risk says that men were about 3 times as likely to die as women.  There's also shortcut way to calculate this number:
```{r}
t1 = xtabs(~sex + survived, data=TitanicSurvival)
relrisk(t1)
```

Note that the `relrisk` function expects the predictor variable to be along the rows and the outcome variable to be along the columns.

### Permutation tests: shuffling the cards

A natural follow-up question is: could this association have arisen due to chance?  One way of addressing this is by something called a permutation test, which explicitly breaks any association between the predictor and the response by "shuffling the cards."  We will build up this idea in stages.

First let's create a simple data set with two variables called "sesame_street" and "army":
```{r}
sesame_street = c('A', 'B', 'C', 'D', 'E')
army = c('alpha', 'bravo', 'charlie', 'delta', 'echo')
data.frame(sesame_street, army)
```
If we look along the rows of this data set, we see a perfect correspondence between the Sesame Street and Army versions of the alphabet.  However, we can break this correspondence by shuffling one set of letters:
```{r}
data.frame(shuffle(sesame_street), army)
```

Try executing the line above a few times. Each time you'll get a different permutation of the Sesame Street alphabet, and therefore a data frame in which the correspondence in the original data set has been broken.

We can do the same thing for the Titanic data.  Try this block a few different times:
```{r}
titanic_shuffle = data.frame(shuffle(TitanicSurvival$sex), TitanicSurvival$survived)
# Look at the first 10 lines
head(titanic_shuffle, 10)
```

In this new data frame `titanic_shuffle`, any connection between sex and survival status has been explicitly broken.  This simple trick of shuffling the cards allows us to assess the plausible range of values for the relative risk under the assumption that there is no association between the two variables.  We can do this a little more concisely as follows:

```{r}
t1_shuffle = xtabs(~shuffle(sex) + survived, data=TitanicSurvival)
relrisk(t1_shuffle)
```

Try executing the code block above a few times.  You notice that each time you calculate the relative risk, you get something much closer to 1 than for the actual data.  Let's use a Monte Carlo simulation to repeat this process 1000 times.  This will allow us to make a histogram that shows the sampling distribution of the relative risk under the null hypothesis of no assocation between sex and survival status.  
```{r}
permtest1 = do(1000)*{
  t1_shuffle = xtabs(~shuffle(sex) + survived, data=TitanicSurvival)
  relrisk(t1_shuffle)
}
head(permtest1)
hist(permtest1$RR)
```

When the cards are shuffled -- and therefore when any possible connection between sex and survival status is explicitly broken -- the relative risk almost never falls outside the interval (0.85, 1.15).  This is pretty convincing evidence that the actual value we observed (about 3) could not have arisen due to chance.


### Using other statistics in the permutation test

One nice feature of the permutation test is that we can use any measure of association we want.  For example, we might prefer to use the [odds ratio](http://en.wikipedia.org/wiki/Odds_ratio) rather than the relative risk.  Let's recall our table of proportions:

```{r}
t1 = xtabs(~sex + survived, data=TitanicSurvival)
p1 = prop.table(t1, margin=1)
p1
```

Now we can use the propotions to calculate the odds ratio by hand.
```{r}
odds_female = p1[1,1]/p1[1,2]
odds_male = p1[2,1]/p1[2,2]
odds_male/odds_female
```

As with the relative risk, there is a shortcut to calculating the odds ratio:
```{r}
t1 = xtabs(~sex + survived, data=TitanicSurvival)
oddsRatio(t1)
```

Let's now use the odds ratio in a permutation test.
```{r}
permtest2 = do(1000)*{
  t1_shuffle = xtabs(~shuffle(sex) + survived, data=TitanicSurvival)
  oddsRatio(t1_shuffle)
}
head(permtest2)
hist(permtest2$OR)
```

This histogram makes it clear that the actual odds ratio of 11.3 far exceeds what we'd expect under the hypothesis that there is no association between sex and survival status.


### The chi-squared statistic

Finally, another common measure of association is the [chi-squared statistic](http://en.wikipedia.org/wiki/Pearson's_chi-squared_test#Calculating_the_test-statistic).  This statistic compares the counts in each cell of the table to what would be expected under the hypothesis that the rows and columns are independent of each other.
```{r}
t1 = xtabs(~sex + survived, data=TitanicSurvival)
chisq(t1)
```

We'll encounter this statistic again when we look at more complicated contingency tables.  The relative risk and odds ratio make sense for 2x2 tables, but the chi-squared statistic generalizes more readily.

